---
title: "Alz random forest"
author: "Alex"
date: "2024"
output: html_document
---


```{r setup,include=FALSE,results=FALSE,warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = 'hold')
#Load Libraries and functions Dr. Abdoli made
library(magrittr)
library(dplyr)
library(tidyr)
library(tidyselect)
library(kableExtra)
library(NACCdata)
library(glmnet)
library(car)
library(tibble)
library(ggmosaic)
library(ROCit)
library(pROC)
library(caret)
library(randomForest)
library(dplyr)
library(rpart.plot)
library(plyr)


colTrim <- function(x, width){
  x <- trimws(x)
  sapply(x, strtrim, width=width)
}
source('describeDS.R')
source('briefSummary.R')
source('xval.R')
```

### Get Data:  
These are the variables that will be used in the model. The order of variables are chosen for convenience of comparing coefficients; the order should not significantly affect the overall results.

```{r}
col2clean <- c('NACCID', 'NACCALZD', 'NACCAGE', 'BPDIAS', 'BPSYS',
               'HXHYPER','HYPERCHO', 'HXSTROKE','CVHATT', 'CVCHF',
               'CVAFIB','NACCBMI', 'DIABETES', 'B12DEF', 'DEP2YRS', 
               'SMOKYRS', 'TOBAC100', 'PACKSPER', 'ALCOHOL', 
               'NACCNIHR', 'HISPANIC','EDUC', 'SEX', 'MARISTAT',
               'NACCVNUM', 'NACCNIHR_updated')
cat(sprintf('{%s}', paste(col2clean, collapse = ', ')), sep='\n')
```

### Variable Specification
The variable cleaning specifications are shown below:

```{r}
#reading the cleaning instructions
glmDS0 <- read.csv('cleanUDS_EDA.csv')

#making table
glmDS0 %>% dplyr::select(-c(Rename, Comment, MissingValComment)) %>% 
  dplyr::filter(GLM!="", Name %in% col2clean) %>% 
  mutate(CurrentVal=colTrim(CurrentVal, 20)) %>% 
  mutate(NewLevel=colTrim(NewLevel, 20)) %>% 
  kbl(caption = 'UDS Variables for the Study') %>% 
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "hover", "condensed",
                                      "responsive")) %>% 
  column_spec(2, width_max = '2in') %>% 
  column_spec(3, width_max = '2in') %>% 
  scroll_box(height = "400px")
```

# Create the data frame
```{r}
#make data frame
X <- NACCdata::cleanDS(
  cleanDF = glmDS0 %>% filter(GLM!="", Name %in% col2clean)
    ) 

#checking race statements
#how many people are multiracial and Hispanic
nrow(X %>% 
  filter(NACCNIHR == "Multiracial" & HISPANIC == "Hispanic"))
#how many people are any race and hispanic? 
nrow(X %>% 
  filter(NACCNIHR != "" & HISPANIC == "Hispanic"))
#how many people have a NA race value and are non hispanic
nrow(X %>% 
  filter(is.na(NACCNIHR) & HISPANIC == "non-Hispanic"))
#how many people  have a NA race value and are hispanic
nrow(X %>% 
  filter(is.na(NACCNIHR) & HISPANIC == "Hispanic"))

#make updated race variable
X <- X %>% 
  mutate(NACCNIHR_updated = case_when(
    NACCNIHR != "" & HISPANIC == "Hispanic" ~ "Multiracial",
    is.na(NACCNIHR) & HISPANIC == "non-Hispanic" ~ "Other",
    is.na(NACCNIHR) & HISPANIC == "Hispanic" ~ "Hispanic",
    TRUE ~ NACCNIHR
  ))

#checking to see if correct
table(X$NACCNIHR_updated)


#set reference category
X$NACCNIHR_updated <- relevel(factor(X$NACCNIHR_updated), ref = "White")

#how many records
X <- X %>% filter(NACCVNUM==1)
cat(sprintf('%d records is read.\n', dim(X)[1]))
```

## Incomplete Cases {.tabset}
The incomplete cases are dropped and the counts of NA's before and after are reported. Then, variables are summarized in a brief format more appropriate for publications. 

### Counting NAs
```{r}
cat('Before removing incomplete observations:\n\n')
(naCounts <- sapply(colnames(X), \(v) sum(is.na(X[[v]]))))

X <- X %>% 
  drop_na(ALCOHOL) %>% 
  drop_na(EDUC) %>% 
  drop_na(MARISTAT) %>% 
  drop_na(NACCBMI) %>% 
  drop_na(PACKSPER) %>% 
  drop_na(B12DEF) %>% 
  drop_na(BPDIAS) %>% 
  drop_na(BPSYS) %>% 
  drop_na(CVAFIB) %>% 
  drop_na(CVCHF) %>% 
  drop_na(CVHATT) %>% 
  drop_na(DEP2YRS) %>% 
  drop_na(DIABETES) %>% 
  drop_na(HXHYPER) %>% 
  drop_na(HXSTROKE) %>% 
  drop_na(HYPERCHO) %>% 
  drop_na(NACCNIHR_updated)

cat('\n\n\nAfter removing incomplete observations:\n\n')
sapply(colnames(X), \(v) sum(is.na(X[[v]])))

#mtry= 15
#minn= 100
#tree 1000
```

## Only AD and Normal Cognition
```{r}
uds <- X %>% mutate(NACCALZD=factor(NACCALZD, c("Normal", "Alzheimer"))) %>%
  filter(!is.na(NACCALZD))

nrow(uds)
```

```{r Biomarker data cleaning and join the table with uds and biomarker data}
library(dplyr)
library(plyr)

bio <- NACCdata::biomarker_data %>%
  dplyr::select(c(NACCID, CSFABETA, CSFTTAU))

u_b <- left_join(uds, bio, by = "NACCID") %>%
  drop_na(CSFTTAU) %>% 
  drop_na(CSFABETA)
```

```{r}
library(tidymodels)
set.seed(1880)

uds_split <- initial_split(u_b, prop = 0.80) 

uds_train <- training(uds_split)
uds_test <- testing(uds_split)
```


```{r}
treeC_model <- decision_tree(mode = "classification", engine = "rpart", cost_complexity = tune())
```


```{r tree-tidy recipe}
treeC_recipe <- recipe(
  NACCALZD ~ NACCAGE + BPDIAS + BPSYS +HXSTROKE + CVAFIB +NACCBMI + B12DEF + DEP2YRS + PACKSPER + ALCOHOL + EDUC + MARISTAT + NACCNIHR_updated + CSFTTAU + CSFABETA,
  data = uds_train
)

treeC_wflow <- workflow() |>
  add_model(treeC_model) |>
  add_recipe(treeC_recipe)
```


```{r}
set.seed(1332)
uds_kfold <- vfold_cv(uds_train, v = 5, repeats = 3) 

treeC_tune1 <- tune_grid(treeC_model, 
                      treeC_recipe, 
                      resamples = uds_kfold, 
                      metrics = metric_set(mn_log_loss),
                      grid = grid_regular(cost_complexity(range = c(-4, 6)), levels = 20))

autoplot(treeC_tune1)
```


```{r select best treeC}
treeC_best <- select_by_one_std_err(
  treeC_tune1,
  metric = "mn_log_loss",
  desc(cost_complexity)
)
treeC_best
```


```{r fit best model}
treeC_wflow_final <- finalize_workflow(treeC_wflow, parameters = treeC_best) 

treeC_fit <- fit(treeC_wflow_final, data = uds_train)
treeC_fit
```


```{r}
extract_fit_engine(treeC_fit) |>
  plot() 
```


```{r augment treeC fit}
broom::augment(treeC_fit, new_data = uds_test) |>
  conf_mat(truth = NACCALZD, estimate = .pred_class)
```


```{r rfC-tidy model}
rfC_model <- rand_forest(mode = "classification", engine = "ranger") |>
  set_args(seed = 395,
           importance = "permutation",
           mtry = tune()
  )

rfC_recipe <- recipe(
  NACCALZD ~ NACCAGE + BPDIAS + BPSYS +HXSTROKE + CVAFIB + NACCBMI + B12DEF + DEP2YRS + PACKSPER + ALCOHOL + EDUC + MARISTAT + NACCNIHR_updated + CSFTTAU + CSFABETA,
  data = uds_train
)

rfC_wflow <- workflow() |>
  add_model(rfC_model) |>
  add_recipe(rfC_recipe)
```

Now we tune the model:

```{r tune model kfold rfC}
n_predictorsC <- sum(rfC_recipe$var_info$role == "predictor")
manual_gridC <- expand.grid(mtry = seq(1, n_predictorsC))

rfC_tune1 <- tune_grid(rfC_model, 
                      rfC_recipe, 
                      resamples = uds_kfold, 
                      metrics = metric_set(mn_log_loss, accuracy),
                      grid = manual_gridC)

autoplot(rfC_tune1)
```



```{r select best rfC}
rfC_best <- select_best(
  rfC_tune1,
  metric = "mn_log_loss"
)
```



```{r fit rfC model}
rfC_wflow_final <- finalize_workflow(rfC_wflow, parameters = rfC_best) 
rfC_fit <- fit(rfC_wflow_final, data = uds_train)
rfC_fit
```
```{r}
broom::augment(rfC_fit, new_data = uds_test) |>
  conf_mat(truth = NACCALZD, estimate = .pred_class)
```
```{r}
brier_class(NA)
```


```{r rfC OOB Brier Score and vip}
library(vip)
rfC_engine <- rfC_fit |> extract_fit_engine()

rfC_engine |> pluck("prediction.error")

vip(rfC_engine, scale = TRUE)

```


```{r}
NACCAGE + BPDIAS + BPSYS +HXSTROKE + CVAFIB +NACCBMI + B12DEF + DEP2YRS + PACKSPER + ALCOHOL + EDUC + MARISTAT + NACCNIHR_updated 

#From Biomarker
CSFTTAU+ CSFTPAU + CSFABETA
```



